Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-29 Manuscript/MPESubmission/Revision/response.tex

% TG: I'm confused for the second part of the question:
% TG: In figure 4 and in Table 1, if you look at the Bayesian consensus (grey) trees and the Bayesian posterior trees (orange) for RF, they do not really overlap. Ever. The 0.11 probability must come from the parameters combinations that, when is really shitty must show a little bit of overlap. But for each parameter one by one: nope. No overlap.
% TG: On the other side, in figure 4 and Table 1, for Bayesian consensus and Bayesian posterior for Tr, the do overlap much more. And that's what's in the Table 1 (min 0.21; median 0.56; and max 0.87).
% TG: Did the reviewer just miss read the labels of the Table 1 or am I miles away from what he means?

\item{\textcolor{blue}{\textbf{Results - Bayesian Triplets:} Another important result that is not mentioned is that using the wildcard taxon metric (Tr), Bayesian consensus trees do quite poorly with missing data.
Values in Figures 4 and 5 clearly drop below zero, although negative values are not illustrated on the axis, and this indicates that the tree topologies are no better than, or even worse than, random trees.
It is especially true when fossils are missing data, and that is going to be an important finding with practical application but must be addressed. }}

We modified the figures 4 and 5 to show Normalised Triplets metric values below 0 and added a dashed line at 0 to underline the difference in scale between both metrics.
% TG: a a paragraph in the results about the wild-card fossils
As mentioned in the previous comment (reviewer 1 comment 9), this is due to the fact that the Bayesian consensus tree will be less and less resolved when missing data increases.
This has two related effects.
The first one being to retain "good" clade conservation (i.e. avoiding "false-positive" nodes by collapsing unsupported ones).
And the second one is that when these clades are collapsed, it reduces the amount of available triplets and thus increases the "wild-card taxa" score (Normalised Triplets metric).
In fact, when a clade ((a;b);c); is collapsed to (a;b;c), the whole clade is conserved (a, b and c are still in a common clade) however they also all become wild-card taxa: the position of a regarding b and c is now incorrect.

@@

We fixed this graphical mistake as state above in reviewer 1 comment 13.

\item{\textcolor{blue}{\textbf{Table caption:} Table caption should include a line about 'pooling across missing data schemes' to clarify what is presented.}}

\item{\textcolor{blue}{\textbf{Previous research:} The missing-data problem has a long history in the literature and there are references to some of those older and recent papers.
But I feel it is necessary to summarize the following questions: what did those studies find, what are the questions those studies left open, and how does this study address them?}}

% specifically state line by line where we dealt with which and which question.

@@

\item{\textcolor{blue}{\textbf{line 76 and throughout the manuscript:} watch out for inconsistent parentheses.}}

We fixed this issue.

\item{\textcolor{blue}{\textbf{line 76 and throughout:} correct use of the word 'data' from singular to plural throughout}}

%We corrected the plural/singular of data throughout the text. 
% TG: actually I'm not going to change it:
We thank the reviewer for pointing out this inconsistency, however, 'data' is now commonly used for the singular case (Pinker, S. \textit{The Sense of Style: The Thinking Person's Guide to Writing in the 21st Century}. Penguin, 2014. p 271; and Guardian and Observer style guide (2015): \url{http://goo.gl/cmKjNe}).

@@

We modified the previous statement to: "but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but using topological constraints" (line @@@).

\item{\textcolor{blue}{\textbf{lines 80-103:} (1) the authors need to clearly justify the importance of this study compared the many studies that have investigated the effects of missing data on topology inference.
The justification currently given is not particularly strong and the specific unique questions asked in this study that are currently unanswered should be made clear.
The current argument does not always seem justified by the literature review.
For example, line 81 the authors argue that papers by Wiens and Pattinson examined morphological or molecular matrices separately, but in Pattinson, for example, combined datasets for living species were used to simulate extinct species by deleting the molecular and parts of the morphological datasets from living species.
The authors could argue that a limitation of the Pattinson study, which they address here, was the limited scheme for generating missing data from the datasets.
In this study, the authors perturbed the datasets more thoroughly in a clearly structured experimental design to investigate the effects of eliminating living versus fossil data or all data overall.
That is one aspect of this study that I perceived as new and unique.
(2) Another aspect of this study design that could be new but is not explored here is the effect of the missing data on the morphological substitution rate parameter estimates.
Can you recover the true values from the perturbed datasets?
This study also used a unique way of evaluating the tree inference under missing data scenarios.
These and other aspects of the study uniqueness must be clearer.}}

(1) - We thank the reviewer for this judicious comment and emphasized on both the novelty aspect of our study and on the questions we are actually testing as follow:

"Until now, only few attempts have been made to study study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using combined molecular and morphological data; e.g. Wiens et al., 2005; Pattinson et al., 2014). For example, Wiens et al. (2005) only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data. Furthermore, Pattinson et al. (2014) assess the effect of removing both molecular and morphological data (“artificial extinction”) on the topology. These studies, however, are limited by their empirical-only approach (Wiens et al., 2005) and the scheme of generating missing data (Pattinson et al., 2014)."

"In this study, we propose a theoretical assessment of the effect of missing data in the Total Evidence method by thoroughly eliminating living or fossil data or all data overall. We test the effect of missing data by measuring two crucial aspects of topology in both Maximum Likelihood and Bayesian Inference methods: (i) the conservation of clades and (ii) the displacement of wild-card taxa."
%Pattinson: artificial fossil - removing molecular data from living species to see how topology changes. Pattinson use combined analysis but only to understand the patterns of fossilisation (via artificial fossil).
%Wiens only focuses on a special case
%Limitation of Pattinson was the limited scheme for generating missing data from the datasets (specific to primates and observed patterns).
%In our case, we (a) don't focus on any group or any pattern, our approach is purely theoretical.
%The difference from Wiens or Pattinson is that we thoroughly investigate the effects of eliminating living versus fossil data or all data overall.
%(b) We also propose a thorough comparison between the ML and Bayesian inference and a measure two clear effects that missing data could have on topology: displacing wild-card taxa and not conserving clades structure.

@@

%Emphasize on the questions and the uniqueness
%The authors could argue that a limitation of the Pattinson study, which they address here, was the limited scheme for generating missing data from the datasets.
%In this study, the authors perturbed the datasets more thoroughly in a clearly structured experimental design to investigate the effects of eliminating living versus fossil data or all data overall.
%That is one aspect of this study that I perceived as new and unique.
%This study also used a unique way of evaluating the tree inference under missing data scenarios.

(2) - the second point underlined by the reviewer in this specific comment strikes use as really interesting. However, we considered this study to be already long and complex. We therefore think that adding an extra question to this study (i.e. what is the effect of missing data on the morphological substitution rate parameter estimates?) might loose the reader.
Nonetheless, we added the result of such analysis in the \textbf{Appendix-A Morphological rates estimations results} section.
We linked to this analysis in the methods part as follows:
"The detailed MrBayes parameters are available in Appendix A along with details on the $\alpha$ parameter estimation."

% TG: what do you think? I don't want to attract the reviewer's attention too much (i.e. that's another shit load of analysis to do to do it properly) and certainly not the reader's attention either (it's not relevant to the whole paper) but sticking that to the supplementary kind of does what the reviewers wants to know.

@@

\item{\textcolor{blue}{\textbf{line 185 to 188:} streamline, e.g. '…sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A).'}}

We modified the following sentence as suggested: "sampling with a probability of 0.85 for two states characters and 0.15 for three state characters (based on an empirical review of published matrices, see Appendix A and Fig. A1 within)." (line @@@).

\item{\textcolor{blue}{\textbf{line 193 to 196:} again, justify why gamma with shape of 0.5 is appropriate for minimizing homoplasy. This distribution places highest probability on substation rates less than 0.5, when one of the cited references potentially suggests that the lowest topological error was found at rates of ~0.5-1.5, and that very low substitution rates incurred error similar to very high rates ((Wright and Hillis 2014) 2014 Figure 3). Other cited references do not clearly state that this distribution and shape setting is optimal, so the authors really must justify. }}

We initially chose this parameter from personal communication with April Wright and to be consistent with the molecular rate parameter.
We like to point out that, in figure 3 from Wright and Hillis 2014, even if the Topological error is high when rates are near 0, it is mainly the case for parsimony and both parsimony and Mk with missing data, the fitted model for the Mk model with no missing data (light purple line) is still at lowest when the rates are near 0.
Topological error, however, becomes minimal for all scenari (Parsimony and Mk, with or without missing data) at a rate of 0.5 (see modified figure 3 from Wright and Hillis (2014) in this response; we added the tick marks and the 0.5 rate limit by using the ruler in Adobe Illustrator).

\begin{figure}[]
\centering
    \includegraphics[keepaspectratio=true]{response_fig1.pdf}
\caption{Figure 3 from Wright and Hillis 2014 with added x axis ticks. The red line represents the 0.5 rate.}
\end{figure}

We do agree, however, that if rates get lower than 0.5, variance in topological error increases (especially for the Mk model with missing data).
Therefore, we implemented the \textit{a posteriori} matrix selection step to our analysis (i.e. keeping only the ``complete'' matrices that led to trees with a median bootstrap $>$ 50; explained lines 259-268 in the submitted manuscript and lines @@@ and @@@ in the revised one).

We rephrased the section underlined by the reviewer as follows:

"We used low evolutionary rate parameters to be consistent with the molecular rates parameters and to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal (Wright and Hillis, 2014). Note that Wright and Hillis (2014) have shown that low morphological rates (< 0.5) increases variance in topological error but we discarded simulations with such topological error by selecting only matrices with a “fair” phylogenetic signal (see Estimating phylogenies section below; Zander, 2004)."
